Penalized regression for left-truncated and right-censored survival data

### 2021-01-07 manuscript.tex
update discussion and follow-up plot

\begin{abstract}
    High-dimensional data are becoming increasingly common in the medical field as large volumes of patient information are collected and processed by high-thoroughput screening, electronic health records (EHRs), and comprehensive genomic testing. Statistical models that attempt to study the effects of many predictors on survival typically implement feature selection or penalized methods to mitigate the undesirable consequences of overfitting. In some cases survival data is also left-truncated which can create an immortal time bias, but penalized survival methods that adjust for left truncation are unavailable in common software. To address these challenges, we implement a penalized Cox proportional hazards model for left-truncated and right-censored survival data and assess implications of left truncation adjustment on bias and inference. We use simulation studies and examples from high-dimensional real-world EHR and genomics data to highlight the pitfalls of failing to account for left truncation in survival modeling. The extension of survival modeling to left-truncated data is available in the latest release of \pkg{glmnet}.
\end{abstract}

@@

Survival is defined as the time from diagnosis until death, but the study does not begin until 2011 (dashed blue vertical line). Patients 2 and 5 (in red) are excluded from the analysis because of death and right censoring, respectively, prior to study eligibility. A naive analysis on this data would therefore be biased because it only contains survival time for ``super-survivors'' (those who have survived a long time, such as patients 1 and 3) and the recently-diagnosed.

@@

In our real-world use case, an analyst could have mistakenly concluded that the high C-index in the non-adjusted model was indicative of good model performance when it was in fact driven primarily by a high correlation between predictors and left truncation. For instance, in our large $p$ model the hazard ratio for the predictor "year of diagnosis" moved from being the largest in absolute value to $1$ after adjusting for left truncation. The high C-index in the non-adjusted model was artificial because patients diagnosed earlier appeared to survive longer due to immortal time bias.

@@

We conclude with several recommendations. When making predictions with survival data is of interest, it is critical to define the risk set appropriately. To do this requires recognizing the underlying process by which individuals are captured in the data in the first place; i.e., when and how individuals came to be included in the dataset of interest. One of the motivations for the current work derives from the advent of biobank and genomic test databases that store information on patient illness - presumably after the patient has been diagnosed and ``at risk'' for some time. In these data, overlooking the ``delayed entry'' of at-risk patients could result in misleading conclusions, as we demonstrate in this work. Additionally, patient cohorts marked by a milestone entry, such as genomic testing, may not only be subject to immortal time bias but also to a temporal selection bias. The latter may occur if left truncation time is correlated with survival time, for example, if genomic testing is ordered selectively for patients with worse prognosis. This was uncovered in a cohort of multi-stage cancer patients who received genomic testing \citep{kehl2020selectionbias} and can affect inference even in the presence of a left truncation adjustment. While the cohort of lung cancer patients in our real-world use case was restricted to stage IV diagnoses, mitigating this potential selection bias \citep{kehl2020selectionbias}, it can be adjusted for by modeling the association between left truncation time and survival \citep{thiebaut2004choice, chiou2019transformation}. Finally, we recommend that models should be evaluated using both discrimination and calibration based metrics.

The implementation of this approach is available in \pkg{glmnet} v.X.X.X on CRAN.

>>>>

\begin{abstract}
    High-dimensional data are becoming increasingly common in the medical field as large volumes of patient information are collected and processed by high-thoroughput screening, electronic health records (EHRs), and comprehensive genomic testing. Statistical models that attempt to study the effects of many predictors on survival typically implement feature selection or penalized methods to mitigate the undesirable consequences of overfitting. In some cases survival data is also left-truncated which can create an immortal time bias, but penalized survival methods that adjust for left truncation are not commonly implemented. To address these challenges, we implement a penalized Cox proportional hazards model for left-truncated and right-censored survival data and assess implications of left truncation adjustment on bias and inference. We use simulation studies and examples from high-dimensional real-world EHR and genomics data to highlight the pitfalls of failing to account for left truncation in survival modeling.
\end{abstract}

@@

Survival is defined as the time from diagnosis until death, but the study does not begin until 2011 (dashed blue vertical line). Patients 1 and 2 (in red) are excluded from the analysis because of death and right censoring, respectively, prior to study eligibility. A naive analysis on this data would therefore be biased because it only contains survival time for ``super-survivors'' (those who have survived a long time, such as patients 3 and 7) and the recently-diagnosed.

@@

In our real-world use case, an analyst could have mistakenly concluded that the high C-index in the non-adjusted model was indicative of good model performance when it was in fact driven primarily by a high correlation between predictors and left truncation. For instance, in our large $p$ model the hazard ratio for the predictor "year of diagnosis" moved from being the largest in absolute value to $1$ after adjusting for left truncation. The high C-index in the non-adjusted model was artificial because patients diagnosed earlier appeared to survive longer due to immortal time bias. To facilitate model interpretation, we recommend that models should be evaluated using both discrimination and calibration based metrics.

@@

When making predictions with survival data is of interest, it is critical to define the risk set appropriately, as our work demonstrates. To do this requires recognizing the underlying process by which individuals are captured in the data in the first place; i.e., when and how individuals came to be included in the dataset of interest. One of the motivations for the current work derives from the advent of biobank and genomic test databases that store information on patient illness - presumably after the patient has been diagnosed and ``at risk'' for some time. In these data, overlooking the ``delayed entry'' of at-risk patients could result in misleading conclusions, as we demonstrate in this work. Additionally, patient cohorts marked by a milestone entry, such as genomic testing, may not only be subject to immortal time bias but also to a temporal selection bias. The latter may occur if left truncation time is correlated with survival time, for example, if genomic testing is ordered selectively for patients with worse prognosis. This was uncovered in a cohort of multi-stage cancer patients who received genomic testing \citep{kehl2020selectionbias} and can affect inference even in the presence of a left truncation adjustment. While the cohort of lung cancer patients in our real-world use case was restricted to stage IV diagnoses, mitigating this potential selection bias \citep{kehl2020selectionbias}, it can be adjusted for by modeling the association between left truncation time and survival \citep{thiebaut2004choice, chiou2019transformation}.
